---
title: 'Assignment 3 Question 1 MAST90125: Bayesian Statistical Learning'
header-includes:
   - \usepackage{bm}
   - \usepackage{amsmath}
output: 
  pdf_document:
    number_sections: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Due: Friday 25 October 2019**  
\vspace{5 mm}

**There are places in this assignment where R code will be required. Therefore set the random seed so assignment is reproducible.**

```{r}
set.seed(234567)  #Please change random seed to your student id number.
```

## Question One (17 marks)

To explore some properties of Expectation propagation and Hamiltonian Monte Carlo, consider the dataset \texttt{Warpbreaks.csv}, which is on LMS and previously analysed in assignment 2. This dataset contains information of the number of breaks in a consignment of wool. In addition, Wool type (A or B) and tension level (L, M or H) was recorded. As the observed data consists of integer counts, it was assumed that a Poisson distribution should be used to model counts. The probability mass function of a Poisson distribution is

\[ Pr(y_i|\lambda_i) = \frac{\lambda_i^{y_i}e^{-\lambda_i}}{y_i!}.  \]


a) Assuming that the canonical link for observation $i$ can be represented as ${\bf X}_i\bm \beta$, determine the following:

\begin{itemize}
\item The likelihood, $p({\bf y}|\bm \beta)$ and log-likelihood.
\item The first derivative of the log-likelihood with respect to $\bm \beta$.
\end{itemize}



b) If you wish to construct a Bayesian analogue to Poisson regression, what prior(s) would you use?



c) Fit a Poisson regression to the warpbreak data, with Wool type and tension treated as factors, using Hamiltonian Monte Carlo. To ensure identifiability, make Wool type A and tension type H the reference category. You are expected to code this in \texttt{R}, as opposed to fitting the model using \texttt{Stan}. Consider the following values for the number of leapfrog steps $L = 2, 3, 4$. Assume the momentum variable $\phi$ is drawn from a multivariate normal distribution with zero mean and variance-covariance matrix $5{\bf X}'{\bf X}$. Run a single chain for each choice of $L$ for 10000 iterations, and remove 30 \% of iterations as burn-in. Report the following.

\begin{itemize}
\item The posterior mean, standard deviations and 90 \% credible intervals for all parameters, combining the results for all chains. Interpret the 90 \% credible interval.
\item The acceptance rate for each choice of $L$.
\end{itemize}

d) Check each chain obtained converged to the same distribution. For each chain and parameter, create acf plots. Based on this, what do you think was the best choice for $L$?

e) Fit the same model using an expectation propagation algorithm. Report the approximate posterior means, and 90 \% credible interval. Comparing the results obtained using expectation propagation to Hamiltonian Monte Carlo, what \lq bias\rq\hspace{1 mm} do you think has been caused by using approximate inference. 
